---
format:
  revealjs:
    height: 900
    width: 1600
    logo: "images/osu.png"
    slide-number: true
    scrollable: true
    code-annotations: hover
---

## Software: Quick Demo 

1. Load Dependencies & Package

2. Data Preparation

3. Model Configuration

4. Model Fitting

5. MCMC Convergence Diagnostics 

6. Prediction & Evaluation


## 1. Load Dependencies & Package

```{r}
#| eval: false
#| echo: true
#| output: false

library(tidyverse) # Data manipulation, visualization & more
library(devtools) # Package development
devtools::load_all() # Load MTD package in local or HPC environment
```

## 2. Data Preparation

```{r}
#| eval: false
#| echo: true
#| output: false

data <- readRDS("MWH.rds") # Load data

test_start <- 7029

train_dat <- data$WHuv50[1:(test_start - 1)]
test_dat <- data$WHuv50[test_start:length(data$WHuv50)]
```

## 3. Model Configuration

```{r}
#| eval: false
#| echo: true
#| output: false

# Model specification
mtdorder <- 5

# MCMC configuration
mcmc_param <- list(niter = 85000, nburn = 5000, nthin = 20, nreport = 40)

# Bayesian priors
prior <- list(
  u_alp = u_alp,
  v_alp = 1, # hyperparmeter for alpha
  u_be = u_be,
  v_be = 1, # hyperparmeter for beta
  dir_shape = rep(1 / mtdorder, mtdorder), # hyperparmeter for DIR weights
  alpha = 1,
  alpha_0 = 5,
  a_G0 = 1,
  b_G0 = 3
) # hyperparmeter for SB and CDP weights

# Initialization & sampler tuning
starting <- list(rho = rep(0, mtdorder), alp = u_alp, be = u_be)

tuning <- list(
  step_size = 0.01,
  upper_alp = u_alp + 5, # upper for alpha
  upper_be = u_be + 5
) # upper for beta
```

## 4. Model Fitting

```{r}
#| eval: false
#| echo: true
#| output: false

set.seed(42) # Ensure reproducibility
runtime <- system.time(
  res <- tsMTD(
    obs = train_dat,
    mtdorder = mtdorder, # MTD order
    family = "gamma", # Gamma marginal
    cop_family = 1, # Gaussian copula
    weight = "cdp", # CDP weights
    prior = prior,
    starting = starting,
    tuning = tuning,
    mcmc_param = mcmc_param
  )
)
```

## Results [^hpc]

```{r}
#| eval: false
#| echo: true
#| output: false

saveRDS(res, file = "res_wind_chain1.RData")
saveRDS(runtime, file = "runtime_wind_chain1.RData")
```

[^hpc]: Model fitting was performed on an HPC cluster, and the fitted results and runtime were returned for local analysis.

## 5. MCMC Convergence Diagnostics [^local]

MCMC convergence was assessed across four chains using trace, density, ACF, and Gelman–Rubin diagnostics. 

[^local]: MCMC convergence diagnostics were performed locally using the `coda` R package. 

## 6. Prediction & Evaluation [^demo]

```{r}
#| eval: false
#| echo: true
#| output: false

# Prepare test data
res <- readRDS(file.path(data_dir, "res_wind_chain1.RData"))
test_start <- 7029 - mtdorder
test_dat <- data$WHuv50[test_start:length(data$WHuv50)]

# Construct lag matrix
lags <- lagmat(test_dat, mtdorder)

# Observed values corresponding to predictions
obs <- test_dat[-(1:mtdorder)]

# Generate 95% predictive intervals
probs <- c(0.025, 0.975)

set.seed(42) # Ensure reproducibility
res_pred <- predMTD(
  res = res,
  family = "gamma", # Gamma marginal
  lags = lags,
  probs = probs
)

# Prediction results
pred <- data.frame(
  obs = obs,
  q025 = res_pred[1, ],
  q975 = res_pred[2, ]
)

# Evaluate empirical coverage
covered <- pred$obs >= pred$q025 & pred$obs <= pred$q975
coverage <- mean(covered)
```

[^demo]: Simplified here for demonstration; implemented as reusable functions in the workflow.  