---
format:
  revealjs:
    logo: osu.png
    slide-number: true
    scrollable: true
---

### Copula-Based Mixture Transition Distribution Models for Forecasting Skewed and Zero-Inflated Time Series: Methodology and Comparisons with Deep Learning LSTM Networks  

Frances Lin

Dissertation Defense

Oregon State University

November 5, 2025

## Motivation {.smaller}

Modeling complex patterns in sequence data is a central task across domains such as energy, insurance, and transportation. 

- Real-world time series often show skewness and zero inflation, which can hinder prediction if ignored. 

. . .    

Recent AI advances, such as Recurrent Neural Networks (RNNs) and Long Short-Term Memory (LSTM) networks, effectively capture nonlinear dependence and long-range correlation. 

- Prior work found comparable predictive performance between probabilistic Mixture Transition Distribution (MTD) model and deep learning LSTM for disease spread. 

- Claims of LSTM superiority can be misleading when compared to inappropriate benchmarks. 

## Objectives {.smaller}

Part I and II

> Develop the copula-based Gamma Mixture Transition Distribution (Gamma MTD) model and its zero-inflated extension (ZIGamma MTD) to capture nonlinear dependence, skewed distributions, and semicontinuous patterns. 

. . .

Part III

> Evaluate these models alongside deep learning approaches, specifically the Long Short-Term Memory (LSTM) networks, to assess predictive performance and robustness. 

## Roadmap {.smaller}

Part I Models for Forecasting Skewed Time Series 

- Recap of Previous Work

- Sensitivity Analysis (Skip)

- Coverage Assessment (Skip)

Part II Models for Forecasting Zero-Inflated Skewed Time Series

- Introduction

- Background

- Proposed Method

- Simulation Studies

- Prediction

## Roadmap {.smaller}

Part III Copula-Based Markov MTD Models vs. Deep Learning LSTM Networks

- Introduction

- Background

- Hyperparameter Tuning, Training, and Metrics 

- Simulation Studies 

- Data Application

# Part I Models for Forecasting Skewed Time Series {background-color="#D73F09" color="white"} 

## Introduction {.smaller}

Time series model captures how past values contribute to current value and uses this information to predict future value.

. . .

The mixture transition distribution (MTD) models 

- Extend the autoregressive (AR) models to accommodate discrete, continuous, and mixed time series.

- Model each conditional distribution as a mixture of transition kernels, with random, stochastic weights.

## Introduction and Objectives {.smaller}

Our work builds upon the architecture of the MTD model introduced by Zheng et al. (2022). 

- Includes various applications, such as the Gaussian, Poisson, negative binomial, and Lomax regression MTD models, extending beyond linear, Gaussian dynamics. 

. . .

However, for certain invariant marginal distributions, the transition kernel may either require careful construction or can result in a form that is not explicitly defined or too complex. 

. . .

> **Objectives:** Propose to incorporate copulas into the transition kernels to address this limitation.

## Background 

## Model Framework: DAG; also known as Bayesian network {.smaller}

Let $\{ X_t: t \in \mathbb{N} \}$ be a time series. Construct the model on a DAG, the joint distribution of $X_1$ to $X_t$ can be factorized into a product of conditional distributions as
```{=latex}
\begin{equation}
f(\textbf{x}) = f(x_1) \prod_{t=2}^t f(x_t | \textbf{x}^{t-1}). \tag{1}
\label{eq:joint_distribution}
\end{equation}
```
\scriptsize $f(x_t | \textbf{x}^{t-1})$ is the conditional probability density function (pdf) of current value $X_t$ given all of its past values $\textbf{X}^{t-1} = \textbf{x}^{t-1}$, where $\textbf{X}^{t-1} = \{X_i: i \leq t - 1 \}$ and $\textbf{x}^{t-1} = \{x_i: i \leq t - 1 \}$.

. . .

Each conditional in (1) is modeled as a mixture of transition kernels, with mixture weights, for each one of a given number of lags. 

## Model Framework: Transition kernels & mixture weights {.smaller}

For $t > L$, the MTD model specifies the conditional distribution of $X_t$ given $\textbf{X}^{t-1} = \textbf{x}^{t-1}$ as 
```{=latex}
\begin{equation}
f(x_t | \textbf{x}^{t-1}) = \sum_{l=1}^L w_l f_l (x_t | x_{t-l}). \tag{2}
\label{eq:cond_distribution}
\end{equation}
```
\scriptsize $f_l (x_t | x_{t-l})$ is the conditional pdf of $X_t$ with respect to the $l$th transition kernel given that ${X}_{t-l} = {x}_{t-l}$. $w_l$ are weight parameters, where $w_l \geq 0$ such that $\sum_{l=1}^L w_l = 1$.

. . .

\normalsize

There are two components in (2):

  - Transition kernel, $f_l$, captures the influence of the $l$th lag value on the current value.  
  
  - Weight parameter, $w_l$, determines the relative contribution of that influence. 

## Existing Method to Proposed Method {.smaller}

Based on Proposition 1 by Zheng et al. (2022), the bivariate distribution approach, for example, identifies a bivariate distribution of $(U_l, V_l)$ such that the marginal densities, $f_{U_l}$ and $f_{V_l}$, are equal to a pre-specified stationary marginal density $f_X$ for all $l$ transition kernels. 

. . .

This is facilitated by the use of a copula, which separates the marginal behavior of the random variables from their dependence structure. 

$$
\textbf{transition kernel} \longrightarrow \textbf{copula} \times \textbf{marginal} 
$$
$$
\ f_l (x_t | x_{t-l}) \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ c_l (x_t, x_{t-l}) \ \ \ \ f_X (x_t)
$$

## Copula {.smaller}

Copula is widely used for dependence modeling.

  - Is applied in quantitative finance, reliability engineering, hydrology, etc.

  - Constitutes a substantial research area.

. . .

Copula is a multivariate cumulative distribution function (cdf) where its marginal distribution of each random variables is $Unif(0, 1)$. 

Using copula, any joint distribution, $F$, can be decomposed into two parts: the copula, $C$, and the marginal distributions, $F_j$, $j = 1,..., p$.

  - Is rooted in Sklar’s theorem. 
  
  - Consists of families of copula. 

## Proposed Method {.smaller} 

Replace $f_{U_l}$ and $f_{V_l}$ with a pre-specified stationary marginal density, $f_X$, for every $x_t$ and for all $l$. For $t > L$, the proposed copula-based MTD model specifies the conditional distribution as
```{=latex}
\begin{equation}
f(x_t | \textbf{x}^{t-1}) = \sum_{l=1}^L w_l c_l (x_t, x_{t-l}) f_X(x_t). \tag{3}
\label{eq:cond_distribution_copula}
\end{equation}
```
\scriptsize $c_l (x_t, x_{t-l})$ is the copula density evaluated at $x_t$ and $x_{t-l}$, and $f_X(x_t)$ is the stationary marginal density evaluated at $x_t$. 

. . .

\normalsize

Compared to (2), the transition kernel, $f_l$, is now replaced by two components: 

  - Copula density, $c_l$, captures and controls the strength of the dependence through a dependence parameter.
  
  - Stationary marginal density, $f_X$, describes the marginal behavior through distribution parameters.

## Background: Bayesian Implementation {.smaller} 

Earlier MTD models were built through frequentist approaches. 

Estimation and prediction in the MTD model by Zheng et al. (2022) is constructed with Bayesian methodologies.

  - Our proposed copula-based MTD model builds upon this model version. 
  
  - We consider the Gaussian copula with Gamma marginals, i.e., Gamma MTD model. 

## Background: Bayesian Implementation {.smaller} 

Inference is facilitated through a set of latent variables ${\{Z_t\}}_{t=L+1}^n$. $Z_t = l$ indicates which $l$th kernel or component the data belongs to. 

The posterior distribution of the parameters, based on the conditional likelihood, is
```{=latex} 
\begin{equation}
\begin{split}
p(\boldsymbol{w}, \boldsymbol{\theta}, {\{z_t\}}_{t=L+1}^n | D_n) \propto \pi_w(\boldsymbol{w}) \prod_{l=1}^L \pi_l(\boldsymbol{\theta}_l) \\ 
\prod_{t=L+1}^n \Biggl\{ f_{z_t} (x_t | x_{t-z_t}, \boldsymbol{\theta}_{z_t}) \sum_{l=1}^L w_l \delta_l(z_t) \Biggl\}.
\end{split}
\end{equation}
```
\scriptsize $\boldsymbol{\theta}$ collects parameters for the transition kernel. $D_n = {\{x_t\}}_{t=L+1}^n$ is the data.

. . .

\normalsize 

With priors specifications, the full simulation-based Bayesian estimation and prediction can be achieved using Markov chain Monte Carlo (MCMC) algorithms. 

## Simulation Studies {.smaller} 

We have assessed accuracy and performance of the proposed model. 

- Convergence Diagnostics ($\boldsymbol{w}$, $\boldsymbol{\rho}$, $\alpha$, $\beta$)

- Weight and Dependence Parameters for Copula ($\boldsymbol{w}$, $\boldsymbol{\rho}$)

- Parameters for Marginal Distribution ($\alpha$, $\beta$)

. . .

Simulation studies demonstrate promising results across various scenarios. 

- Scenario 1 and 2: Default setup; Compatible $\boldsymbol{w}$ and $\boldsymbol{\rho}$

- Scenario 1.3 and 1.4: Incompatible $\boldsymbol{w}$ and $\boldsymbol{\rho}$

- Scenario 1.5 and 1.6: Compatible $\boldsymbol{w}$ and $\boldsymbol{\rho}$; Rarely observable patterns

- Scenario 3-6: Usual case for gamma marginals

- Scenario 7-9:Unusual case involving high skewness

## What's New {.smaller} 

Since the Preliminary Exam, we have also conducted:

- Sensitivity Analysis

- Coverage Assessment

. . .

Results appear reasonable, and the estimates are consistent with the true values, indicating that the model is robust to the choice of prior. 

Most parameters achieve full coverage, indicating that the posterior intervals reliably capture the true parameter values.

# Part II Models for Forecasting Zero-Inflated Skewed Time Series {background-color="#D73F09" color="white"}

## Introduction {.smaller} 

Zero-inflated data are characterized by an excess of zero values and  frequently appear in domains such as healthcare, insurance, environment, and transportation. 

Examples include 

- medical costs, 

- insurance claims, 

- precipitation amounts, and 

- safety measures.

If left unaddressed, zero-inflation undermines model robustness and weakens inference.

## Introduction and Objectives {.smaller}

In Part I, we propose the copula-based Gamma MTD model, which enables flexible dependence modeling and accommodates arbitrary continuous marginals. 

However, 

- it remains limited in handling excessive zeros commonly observed in real-world continuous data. 

- Copula-based approaches face identifiability issues when modeling discrete or mixed marginals.

. . .

> **Objectives:** Propose reconstructing the marginal distribution to account for zero-inflation. 

## Background

## Proposed Method {.smaller} 

To construct zero-inflated Gamma for the marginal distribution, 

- The Gamma distribution is first reparametrized in terms of the mean, $\mu$, and the scale parameter, $\beta$. 

- Zero values are then replaced with non-zero values drawn from a uniform distribution. 

. . .

Specifically,
```{=latex} 
\begin{equation}
0 \leftarrow U_i, \tag{4}
\end{equation}
```
where $U_i$ follows a continuous uniform distribution on $(0, \epsilon)$ with $\epsilon$ is a data-driven paramater representing the smallest observed non-zero values. 

## Proposed Method {.smaller} 

The resulting distribution, denoted as $ZIGamma(\mu, \beta, P, \epsilon)$, is expressed as:
```{=latex} 
\begin{equation}
f(x) = 
\begin{cases} 
Unif(0, \epsilon) & \text{with probability } P \\
ShiftedGamma(\mu, \beta; \epsilon) & \text{with probability } 1-P, 
\end{cases}, \tag{5}
\end{equation}
```
where $\mu$ denotes the mean and $\beta$ the scale parameter of the shifted Gamma distribution, $P \in [0, 1]$ the zero-inflated probability, and $\epsilon > 0$ the threshold parameter. 

. . .

In (5), there are two parts: 

- We have already discussed the uniform distribution. 

- The shifted Gamma distribution, $ShiftedGamma(\mu, \beta; \epsilon)$, is a standard Gamma distribution with mean $\mu$ and scale $\beta$ that is shifted to the right by $\epsilon$, with the support $[\epsilon, \infty)$. 

## dzig {.smaller} 

::: {.panel-tabset}

```{r}
#| echo: false
library(tidyverse)
library(gridExtra)
```

```{r}
#| echo: false
dsgamma <- function(x, mu, beta, eps, logd = FALSE) {
  #(y>=epsilon)*dgamma(y-epsilon,shape=mu/beta,scale=beta)               # Nick's
  ifelse(x >= eps,                                                       # ifelse because x is a vec.
         dgamma(x - eps, shape = mu / beta, scale = beta, log = FALSE), 0)
}

dzig <- function(x, mu, beta, eps, Pi, logd = FALSE) {
  #Pi*dunif(y,0,epsilon) + (1-Pi)*dsgamma(y,mu,beta,epsilon)             # Nick's
  density <- Pi * dunif(x, 0, eps, log = FALSE) + (1-Pi) * dsgamma(x, mu, beta, eps, log = FALSE) 
  if (logd){
    return(log(density))
  } else{
    return(density)
  }
}
```

```{r}
#| echo: false
plot_dzig <- function(mu, beta, eps, Pi){
  x <- seq(0, 25, length.out = 500)
  y <- dzig(x, mu = mu, beta = beta, eps = eps, Pi = Pi)
  df <- data.frame(
    x = x, 
    density = y
  )
  ggplot(df, aes(x = x, y = density)) +
    geom_line(size = 1) + 
    labs(x = bquote("dzig("~ 
                mu == .(mu) * " " ~ 
                beta == .(beta) * " " ~ 
                P == .(Pi) * " " ~ 
                epsilon == .(eps) *")"), 
         caption = "") + 
    theme_bw(base_size = 32) + 
    theme(
      plot.margin = unit(c(1, 1, 1, 1), "cm"),      
      axis.title.x = element_text(size = 32), 
      axis.title.y = element_text(size = 32), 
      axis.text.x = element_text(size = 32), 
      axis.text.y = element_text(size = 32), 
      plot.caption = element_text(size = 32, hjust = 0) 
    ) -> p
  return (p)
}
```

### $P = 0.1$ $\epsilon = 0.1$
```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.1, eps = 0.1)
```

### $P = 0.1$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.1, eps = 0.4)
```

### $P = 0.5$ $\epsilon = 0.1$

```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.5, eps = 0.1)
```

### $P = 0.5$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.5, eps = 0.4)
```

### $P = 0.7$ $\epsilon = 0.1$

```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.7, eps = 0.1)
```

### $P = 0.7$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_dzig(mu = 7, beta = 1, Pi = 0.7, eps = 0.4)
```

:::

## pzig {.smaller} 

::: {.panel-tabset}

```{r}
#| echo: false
psgamma <- function(x,mu,beta,eps) { 
  (x>=eps)*pgamma(x-eps,shape=mu/beta,scale=beta) 
}

pzig <- function(x,mu,beta,eps,Pi) { 
  Pi*punif(x,0,eps) + (1-Pi)*psgamma(x,mu,beta,eps) 
}
```

```{r}
#| echo: false
plot_pzig <- function(mu, beta, eps, Pi){
  x <- seq(0, 25, length.out = 500)
  y <- pzig(x, mu = mu, beta = beta, eps = eps, Pi = Pi)
  df <- data.frame(
    x = x, 
    probability = y
  )
  ggplot(df, aes(x = x, y = probability)) +
    geom_line(size = 1) + 
    labs(x = bquote("pzig("~ 
                    mu == .(mu) * " " ~ 
                    beta == .(beta) * " " ~ 
                    P == .(Pi) * " " ~ 
                    epsilon == .(eps) * ")"), 
         caption = "") + 
    theme_bw(base_size = 32) + 
    theme(
      plot.margin = unit(c(1, 1, 1, 1), "cm"),      
      axis.title.x = element_text(size = 32), 
      axis.title.y = element_text(size = 32), 
      axis.text.x = element_text(size = 32), 
      axis.text.y = element_text(size = 32)
      #plot.caption = element_text(size = 18, hjust = 0) 
    ) -> p
  return (p)
}
```

### $P = 0.1$ $\epsilon = 0.1$
```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.1, eps = 0.1)
```

### $P = 0.1$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.1, eps = 0.4)
```

### $P = 0.5$ $\epsilon = 0.1$

```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.5, eps = 0.1)
```

### $P = 0.5$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.5, eps = 0.4)
```

### $P = 0.7$ $\epsilon = 0.1$

```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.7, eps = 0.1)
```

### $P = 0.7$ $\epsilon = 0.4$

```{r}
#| echo: false
plot_pzig(mu = 7, beta = 1, Pi = 0.7, eps = 0.4)
```

:::

## Simulation Setting {.smaller} 

Simulated data: 

- With weight $\boldsymbol{w}$, dependence for Gaussian copula $\boldsymbol{\rho}$, mean $\mu$, scale $\beta$, zero-inflated probability $P$, and threshold parameter $\epsilon$, we generate $n = 2000$ observations from the copula-based ZIGamma MTD model.

Model fitting: 

- Set the order $L = 5$ and consider the Gaussian copula with Gamma marginals.

MCMC setting: 

- Run the Gibbs sampler for $165, 000$ iterations, discard the first $5000$ iterations as burn-in, and collect samples every $20$ iterations.

- Run four MCMC chains with $8000$ iterations each for all of the following scenarios. 

## Simulation Studies {.smaller} 

We have assessed accuracy and performance of the proposed model.

- Convergence Diagnostics ($\boldsymbol{w}$, $\boldsymbol{\rho}$, $\mu$, $\beta$, $P$, $\epsilon$)

- Weight and Dependence Parameters for Copula ($\boldsymbol{w}$, $\boldsymbol{\rho}$)

- Parameters for Marginal Distribution ($\mu$, $\beta$, $P$, $\epsilon$)

Simulation studies demonstrate promising results across various scenarios. 

- Scenario 1 and 2: Default setup; Compatible $\boldsymbol{w}$ and $\boldsymbol{\rho}$

- Scenario 3-6: Usual case for zero-inflated gamma marginals

- Scenario 7-9: Unusual case involving high skewness

## Simulation Studies {.smaller} 

In all scenarios, we vary $P$ and $\epsilon$, resulting in six cases: 

- $P = 0.1$, $\epsilon = 0.1$

- $P = 0.1$, $\epsilon = 0.4$

- $P = 0.5$, $\epsilon = 0.1$

- $P = 0.5$, $\epsilon = 0.4$

- $P = 0.7$, $\epsilon = 0.1$

- $P = 0.7$, $\epsilon = 0.4$

For brevity, we will only show a few examples.

## Simulation Results {.smaller} 

## Prediction {.smaller} 

# Part III Copula-Based Markov MTD Models vs. Deep Learning LSTM Networks {background-color="#D73F09" color="white"} 

## Introduction {.smaller} 

## Introduction and Objectives {.smaller}

## Background {.smaller} 

## Network Configuration {.smaller} 

## Experimental Setup {.smaller} 

## Simulation Studies {.smaller} 

## Simulation Results {.smaller} 

## Data Application {.smaller} 

## Conclusion 

## Future Work

## Practical Implications

# Thank you!
